Individual and Differential Harm in Redistricting

### 2022-05-17 paper/paper.Rmd
corrections

We next apply our proposed framework to @2022merrill, ongoing VRA litigation in which civil rights have argued that the existing Black majority district packs Black voters, and a second Black opportunity-to-elect district should be drawn.

@@

We will build up estimates of the overall $H(q)$ and $\Delta H_{\dem-\rep}(q)$ from precinct-by-group harm estimates.
Since we assumed that elections are independent of districting plans, we can first average across elections and then across counterfactual plans.
For a random voter $I$ in precinct $j$, we have for every plan $q$ that the probability of a Democrat winning their district is $$
\Pr(\win_{q(I)}(V, q)=\dem\mid I\in j)
= \Pr(S_{q^{-1}(j),V} > \half)
= \Phi\qty(\frac{\gamma(\bar S_{q^{-1}(j)})}{\sigma^2_\mathrm{hist}}),
$$ where $\Phi$ is the standard Normal CDF. To estimate this probability, we simply replace $\Phi$ with the CDF of the $t_{22}$ distribution ($F_{t_22}$), and $\sigma^2_\mathrm{hist}$ with $\hat\sigma^2_\mathrm{hist}$.
Extending one step further, for two plans $q$ and $\tilde q$, we have

@@

```{=tex}
\begin{equation}\label{eq:appharm2}
\Pr(\win_{q(I)}(V, q)=\rep\text{ and }\win_{\tilde q(I)}(V, \tilde q)=\dem\mid I\in j)
= \qty(\Phi\qty(\frac{\gamma(\bar S_{\tilde q^{-1}(j)})}{\sigma^2_\mathrm{hist}})
- \Phi\qty(\frac{\gamma(\bar S_{q^{-1}(j)})}{\sigma^2_\mathrm{hist}}))_+,
\end{equation}
```
where $(\cdot)_+\dfeq \max\{0, \cdot\}$.
If we condition on $V(I)=\dem$, then \eqref{eq:appharm2} is exactly the expected harm to voter $I$ under plans $q$ and $\tilde q$.
The calculation is analogous, up to sign changes, for the expected harm conditioning on $V(I)=\rep$.
This completes the first step: averaging harm within precinct and groups and across elections.

We can then estimate the average across counterfactual plans $\tilde Q$ with the usual Monte Carlo estimator: $$
    \hat h_{j,\dem}(q)
    \dfeq \frac{1}{N_\text{sim}} \sum_{k=1}^{N_\text{sim}}
    \qty(\Phi\qty(\frac{\gamma(\bar S_{\tilde q_k^{-1}(j)})}{\sigma^2_\mathrm{hist}})
        - \Phi\qty(\frac{\gamma(\bar S_{q^{-1}(j)})}{\sigma^2_\mathrm{hist}}))_+,
$$ where $\hat h_{j,\dem}(q)$ is our estimate of the average harm to a Democratic voter in precinct $j$, averaging across $V$ and $\tilde Q$.
The corresponding $\hat h_{j,\rep}(q)$ is computed similarly.

@@

We model the observed vote counts as \begin{align*}
    \votes_{jv} \mid \turn_{jr}, \beta^{(t)}_v
        &\sim \Binlog\qty(\vap_j, \logit(\turn_j^\top\vb p_j) + \beta^{(t)}_v) \\
    \logit(\turn_j) \mid \mu^{(t)}, \Sigma^{(t)}
        &\sim \Norm(\mu^{(t)}, \Sigma^{(t)}) \\
    s_{jv}\votes_{jv} \mid \supp_{jr}, \turn_{jr}, \beta^{(s)}_v
        &\sim \Binlog\qty(\votes_{jv}, \logit\qty(
        \supp_j^\top \frac{\turn_j\circ\vb p_j}{\turn_j^\top\vb p_j}) + \beta^{(s)}_v) \\
    \logit(\supp_j) \mid \mu^{(s)}, \Sigma^{(s)}
        &\sim \Norm(\mu^{(s)}, \Sigma^{(t)}),
\end{align*} where the $\vb*\beta^{(t)}$ and $\vb*\beta^{(s)}$ model election-specific logit shifts, $\Binlog$ indicates the Binomial distribution with the probability paramter specified on the logit scale, and $\circ$ indicates the elementwise (Hadamard) product.
We give $\beta^{(t)}$ an i.i.d. $\Norm(0, 0.6^2)$ prior,and $\beta^{(s)}$ an i.i.d. $\Norm(0, 0.4^2)$ prior, and $\mu^{(t)}$ an i.i.d. $\Norm(0, 1.5^2)$ prior, and $\mu^{(s)}$ an i.i.d. $\Norm(0, 1.0^2)$ prior.
We set priors independently on the variance and correlation components of $\Sigma^{(t)}$ and $\Sigma^{(s)}$; the former received $\Gam(1.5, 1.5/0.5)$ and $\Gam(1.5, 1.5/0.25)$ priors for turnout and support, while the latter received uniform priors (i.e., LKJ priors with shape parameter 1).
Several reparametrizations were made for computational efficiency; the full Stan code for the model is available as part of the replication code.

@@

We model the observed vote counts as \begin{align*}
    \votes_{jv} \mid \turn_{jr}, \beta^{(t)}_v
        &\sim \Binlog\qty(\vap_j, \logit(\turn_j^\top\vb p_j) + \beta^{(t)}_v) \\
    \logit(\turn_j) \mid \mu^{(t)}, \Sigma^{(t)}
        &\sim \Norm(\mu^{(t)}, \Sigma^{(t)}) \\
    \logit(s_{jk}) \mid p_{jk}, \eps &\sim \Norm(p_{jk}, \frac{\eps}{1+\votes_{jv_k}}) \\ 
    p_{jk} &= \beta^{(s)}_k + \alpha_k
        \pref_{j}^\top(\turn_j\circ \vb p_j)\mathrm{diag}(\vb*\gamma)\vb L_k  \\
    \pref_{jq} \mid \Sigma^{(p)} &\sim \Norm(0, \Sigma^{(p)}).
\end{align*} Note that $\pref_j$ is a $q\times r$ matrix of preferences by race and dimension within precinct $j$, and $\turn_j\circ \vb p_j$ is a vector of the percentages of the total VAP in the precinct that belongs to voters of each race, so $\pref_{j}^\top(\turn_j\circ \vb p_j)$ is a vector with $q$ elements containing the preferences by dimension of the precinct's voters, after marginalizing out race.
The matrix $\mathrm{diag}(\vb*\gamma)$ captures the relative scale or importance of each dimension to the overall outcomes.

@@

We give $\beta^{(t)}$ an i.i.d. $\Norm(0, 0.5^2)$ prior,and $\beta^{(s)}$ an i.i.d. $\Norm(0, 4.0^2)$ prior, and $\logit(\mu^{(t)})$ an i.i.d. $\Beta(10, 30)$ prior.
We set priors independently on the variance and correlation components of $\Sigma^{(t)}$ and $\Sigma^{(p)}$; the former has $\Gam(1.5, 1.5/0.25)$ and $\Gam(1.5, 1.5/1.0)$ priors for turnout and support, while the latter have LKJ priors with shape parameters 4.0 and 20.0.
A larger LKJ shape parameter places more weight on correlation matrices closer to diagonal, i.e., it regularizes away from high correlations.
The scale parameters $\vb*\gamma$ have i.i.d. $\Gam(1.5, 1.5/4.0)$ priors, the fudge factors $\vb*\alpha$ have i.i.d. $\Gam(5.0, 5.0)$ priors, and $\eps$ has an $\Expo(1/4.0)$ prior.
Overall, these priors weakly identify the location and scale of the parameters.
Several reparametrizations were made for computational efficiency; the full Stan code for the model is available as part of the replication code.

>>>>

We next apply our proposed framework to @2022merrill, ongoing VRA litigation in which civil rights groups have argued that the existing Black majority district packs Black voters, and so a second Black opportunity-to-elect district should be drawn.

@@

We will build up estimates of the overall $H(q)$ and $\Delta H_{\dem-\rep}(q)$ from precinct-by-group harm estimates.
Since we assumed that elections are independent of districting plans, we can first average across elections and then across counterfactual plans.
For a random voter $I$ in precinct $j$, we have for every plan $q$ that the probability of a Democrat winning their district is $$
\Pr(\win_{q(I)}(V, q)=\dem\mid I\in j)
= \Pr\qty(S_{q^{-1}(j),V} > \half)
= \Phi\qty(\frac{\gamma(\bar S_{q^{-1}(j)})}{\sigma^2_\mathrm{hist}}),
$$ where $\Phi$ is the standard Normal CDF. To estimate this probability, we simply replace $\Phi$ with the CDF of the $t_{22}$ distribution ($F_{t_{22}}$), and $\sigma^2_\mathrm{hist}$ with $\hat\sigma^2_\mathrm{hist}$.
Extending one step further, for two plans $q$ and $\tilde q$, we have

@@

```{=tex}
\begin{equation}\label{eq:appharm2}
\Pr(\win_{q(I)}(V, q)=\rep\text{ and }\win_{\tilde q(I)}(V, \tilde q)=\dem\mid I\in j)
= \max\qty{\Phi\qty(\frac{\gamma(\bar S_{\tilde q^{-1}(j)})}{\sigma^2_\mathrm{hist}})
- \Phi\qty(\frac{\gamma(\bar S_{q^{-1}(j)})}{\sigma^2_\mathrm{hist}}), 0}.
\end{equation}
```
If we condition on $V(I)=\dem$, then \eqref{eq:appharm2} is exactly the expected harm to voter $I$ under plans $q$ and $\tilde q$.
The calculation is the same, up to sign changes, for the expected harm conditioning on $V(I)=\rep$.
This completes the first step: averaging harm within precinct and groups and across elections.

We can then estimate the average across counterfactual plans $\tilde Q$ with the usual Monte Carlo estimator: $$
    \hat h_{j,\dem}(q)
    \dfeq \frac{1}{N_\text{sim}} \sum_{k=1}^{N_\text{sim}}
    \max\qty{F_{t_{22}}\qty(\frac{\gamma(\bar S_{\tilde q_k^{-1}(j)})}{\sigma^2_\mathrm{hist}})
        - F_{t_{22}}\qty(\frac{\gamma(\bar S_{q^{-1}(j)})}{\sigma^2_\mathrm{hist}}), 0},
$$ where $\hat h_{j,\dem}(q)$ is our estimate of the average harm to a Democratic voter in precinct $j$, averaging across $V$ and $\tilde Q$.
The corresponding $\hat h_{j,\rep}(q)$ is computed similarly.

@@

We model the observed turnout $\votes_{jv}$ and vote counts $s_{jk}\votes_{jv}$  as \begin{align*}
    \votes_{jv} \mid \turn_{jr}, \beta^{(t)}_v
        &\sim \Binlog\qty(\vap_j, \logit(\turn_j^\top\vb p_j) + \beta^{(t)}_v) \\
    \logit(\turn_j) \mid \mu^{(t)}, \Sigma^{(t)}
        &\sim \Norm(\mu^{(t)}, \Sigma^{(t)}) \\
    s_{jv}\votes_{jv} \mid \supp_{jr}, \turn_{jr}, \beta^{(s)}_v
        &\sim \Binlog\qty(\votes_{jv}, \logit\qty(
        \supp_j^\top \frac{\turn_j\circ\vb p_j}{\turn_j^\top\vb p_j}) + \beta^{(s)}_v) \\
    \logit(\supp_j) \mid \mu^{(s)}, \Sigma^{(s)}
        &\sim \Norm(\mu^{(s)}, \Sigma^{(t)}),
\end{align*} where the $\vb*\beta^{(t)}$ and $\vb*\beta^{(s)}$ model election-specific logit shifts, $\Binlog$ indicates the Binomial distribution with the probability paramter specified on the logit scale, and $\circ$ indicates the elementwise (Hadamard) product.
We give $\vb*\beta^{(t)}$ an i.i.d. $\Norm(0, 0.6^2)$ prior,and $\vb*\beta^{(s)}$ an i.i.d. $\Norm(0, 0.4^2)$ prior, and $\mu^{(t)}$ an i.i.d. $\Norm(0, 1.5^2)$ prior. 
We set priors independently on the variance and correlation components of $\Sigma^{(t)}$ and $\Sigma^{(s)}$; the former received $\Gam(1.5, 1.5/0.5)$ and $\Gam(1.5, 1.5/0.25)$ priors for turnout and support, while the latter received uniform priors (i.e., LKJ priors with shape parameter 1).
Several reparametrizations were made for computational efficiency; the full Stan code for the model is available as part of the replication code.

The mean support $\mu^{(s)}$ receives an informative prior based on the racially polarized voting estimates of @kuriwaki2021geography, with $\mu^{(s)}\sim\Norm(\logit((0.18, 0.90, 0.35)), (0.04^2, 0.04^2, 0.2^2))$.
These estimates were derived through multilevel regression and poststratification from survey data, calibrated to actual election results.

@@

We model the observed turnout $\votes_{jv}$ and vote shares $s_{jk}$ as \begin{align*}
    \votes_{jv} \mid \turn_{jr}, \beta^{(t)}_v
        &\sim \Binlog\qty(\vap_j, \logit(\turn_j^\top\vb p_j) + \beta^{(t)}_v) \\
    \logit(\turn_j) \mid \mu^{(t)}, \Sigma^{(t)}
        &\sim \Norm(\mu^{(t)}, \Sigma^{(t)}) \\
    \logit(s_{jk}) \mid \pi_{jk}, \eps &\sim \Norm(\pi_{jk}, \frac{\eps}{1+\votes_{jv_k}}) \\ 
    \pi_{jk} &= \beta^{(s)}_k + \alpha_k
        \pref_{j}^\top(\turn_j\circ \vb p_j)\mathrm{diag}(\vb*\rho)\vb L_k  \\
    \pref_{jq} \mid \Sigma^{(p)} &\sim \Norm(0, \Sigma^{(p)}).
\end{align*} Note that $\pref_j$ is a $q\times r$ matrix of preferences by race and dimension within precinct $j$, and $\turn_j\circ \vb p_j$ is a vector of the percentages of the total VAP in the precinct that belongs to voters of each race, so $\pref_{j}^\top(\turn_j\circ \vb p_j)$ is a vector with $q$ elements containing the preferences by dimension of the precinct's voters, after marginalizing out race.
The matrix $\mathrm{diag}(\vb*\rho)$ captures the relative scale or importance of each dimension to the overall outcomes.

@@

We give $\vb*\beta^{(t)}$ an i.i.d. $\Norm(0, 0.5^2)$ prior,and $\vb*\beta^{(s)}$ an i.i.d. $\Norm(0, 4.0^2)$ prior, and $\logit(\mu^{(t)})$ an i.i.d. $\Beta(10, 30)$ prior.
We set priors independently on the variance and correlation components of $\Sigma^{(t)}$ and $\Sigma^{(p)}$; the former has $\Gam(1.5, 1.5/0.25)$ and $\Gam(1.5, 1.5/1.0)$ priors for turnout and support, while the latter have LKJ priors with shape parameters 4.0 and 20.0.
A larger LKJ shape parameter places more weight on correlation matrices closer to diagonal, i.e., it regularizes away from high correlations.
The scale parameters $\vb*\rho$ have i.i.d. $\Gam(1.5, 1.5/4.0)$ priors, the fudge factors $\vb*\alpha$ have i.i.d. $\Gam(5.0, 5.0)$ priors, and $\eps$ has an $\Expo(1/4.0)$ prior.
Overall, these priors weakly identify the location and scale of the parameters.
Several reparametrizations were made for computational efficiency; the full Stan code for the model is available as part of the replication code.
